---
title: "Modelling physical activity in Melbourne"
format: gfm
execute:
  # no data is to be included in this document, unless intentionally over-ridden
  output: false 
knitr:
  opts_chunk: 
    collapse: true
---

This Quarto markdown document is intended to contain code to predict physical activity for sport and recreation excluding walking and cycling, given explanatory variables (age, sex, SES, education, has car, etc) that align with the JIBE synthetic population data.  This will be conducted to predict physical activity for residents of Melbourne, Australia, using the Australian Bureau of Statistics Australian National Health Survey data (2017-18).

## Dependencies

Analysis was conducted using R 4.4.1 using a Quarto markdown document (Quarto 1.5.55) in Positron IDE (2024.10.0), with renv 1.0.11 for package management.

The following packages have been installed using renv:

```
renv::install(c('dplyr','haven','rmarkdown'))
```

In principle, the R environment for this notebook should be able to be restored by running

```
renv::restore()
```

Load libaries

```{r}
library(dplyr)
```

## Data

[National Health Survey](https://www.abs.gov.au/statistics/microdata-tablebuilder/available-microdata-tablebuilder/national-health-survey) (NHS) data for 2017-18 were retrieved from the [Microdata Downloads](https://microdatadownload.abs.gov.au/MicrodataDownload/login.xhtml) section of the Australian Bureau of Statistics website on 3 December 2024.  ABS Microdata were accessed under the [ABS/Universities Australia Agreement (2024)](https://www.abs.gov.au/statistics/microdata-tablebuilder/absuniversities-australia-agreement) by Carl Higgs (RMIT University). The NHS Microdata data descriptions are available for download [here](https://www.abs.gov.au/statistics/microdata-tablebuilder/available-microdata-tablebuilder/national-health-survey#data-item-lists).

NHS microdata are provided in CSV, SAS, SPSS, or Stata formats.  The CSV data do not have labels, hence the `haven` package could be installed to read the labelled data in .dta (Stata) format.  However, perhaps labels are not required --- for now, CSV will be used to keep things simple.


|File (csv, dta, etc) | Description|
|-----|------------|
|NHS17HHB|Household level data|
|NHS17SPB|Person level data|
|NHS17A3B|Alcohol day level data|
|NHS17A4B|Alcohol type level data|
|NHS17CNB|Conditions level data|
|NHS17MDB|Medications level data|
|NHS17HLB|Health Literacy level data|

: ABS NHS 2017-18 Microdata files

The household data contain geographic attributes and could potentially be used to restrict the sample, e.g. to residents of urban areas within Greater Melbourne.  Sensitivity analysis could be conducted to evaluate the impact of this decision, e.g. relative to all persons and all persons living in Australian urban regions.

Household variables of interest include:

| Variable | Description | Comment |
|----------|-------------|-------------------|
| ABSHIDB  | Household identifier | Link with persons |
| GCCSA16  | Greater Capital City Statistical Area (ASGS 2016) | 1 == capital city |
| SOS16    | Sections of State (ASGS 2016) | 0 == Major urban, 1 == Other urban (filter in c(0,1)) |
| STATE16  | State or Territory (ASGS 2016) | 2 == Victoria |
| DWSTQ02  | Dwelling structure | in case of use later |
| HSTENURE | Tenure type of households | in case of use later  |
| SA1SF2DN | SEIFA - Index of Relative Socio-economic Disadvantage - 2016 - SA1 - Deciles - National | |
| GCDISTME | Median commuting distance (kms) |

Note that there are also household geospatial indicators for supermarkets, amenities, fast food an public open space (any, and larger than 1 hecatare) within buffer distances of 400, 1000 and 1500 metres.

Person variables of interest (see data dictionaries for detailed codes) include:

| Variable | Description | Comment |
|----------|-------------|-------------------|
| ABSPID   | Person identifier    | Need to verify that this is unique within households |
| ABSHIDB  | Household identifier | Link with households |
| AGE99    | Age of person | |
| SEX      | Sex of person | 1==Male, 2==Female |
| HIGHLVL  | Level of highest educational attainment | 0==NA, 1==Postgraduate, ... 13==Never attended school |
| EMPSTAT  | Labour force full-time/part-time status | 0==NA, 1==Employed full time ... 6 Not in labour force |
| ... | other variables | more to add |


```{r}
data_dir <- '/Users/E33390/Library/CloudStorage/OneDrive-RMITUniversity/projects/abs/microdata/NHS2017-18_CSV/NHS2017-18_CSV/'
data <- list(
    households = read.csv(paste0(data_dir,'NHS17HHB.csv')),
    persons = read.csv(paste0(data_dir,'NHS17SPB.csv'))
)
```

```{r}
#| output: true

nhs <- left_join(
            data$households,
                # %>% as.data.frame(), 
                # select(
                #     c( 
                #         "GCCSA16", 
                #         "SOS16",
                #         "STATE16", 
                #         "DWSTQ02",
                #         "HSTENURE",
                #         "SA1SF2DN", 
                #         "GCDISTME"
                #     )
                # ) %>% 
                # filter(
                #     STATE16==2 &
                #     GCCSA16==1 &
                #     SOS16 %in% c(0,1)
                # ), # for Victoria GCCSA (i.e. Melbourne) urban regions 
            data$persons %>% 
                select(
                    c(
                        "ABSPID",
                        "ABSHIDB",
                        "AGEB",
                        "SEX",
                        "HIGHLVBC",
                        "EMPTYPE"
                    )
                ), 
        by = "ABSHIDB"
        ) 

nhs %>% summary()
```